Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-09-21 intro.tex
wrote: matrix element factorisation

    {\color{red} Need to write down the Feynman
    rules and draw diagrams. Best to put it in a
    figure that we can refer back to.}

    Note that we have only written down the vertices,
    propagators, and external lines for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

@@

    For a review of PDFs and how they are determined
    see Ref. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs}.
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\dfrac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops (cite here). At leading order (cite here) they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

@@

    From (\ref{eqn:S_prob}), it becomes clear
    that the matrix element is the object which
    relates back to the Lagrangian of the theory
    as it encodes the interactions of the particles.
    We saw in Section \ref{sec:qcd_feynman} that
    interactions could be codified into Feynman
    rules which are pieced together to construct
    Feynman diagrams. In this picture, matrix elements
    are exactly the sum over all Feynman diagrams
    \footnote{technically the interference of
    Feynman diagrams as we are working on the level
    of $|\mathcal{M}|^{2}$}
    for a particular process. However, there are
    an infinite number of Feynman diagrams you can
    draw as there is no constraints on how the
    vertices and propagators can be connected.
    Fortunately because $\alpha_{s}$
    is small in high energy collisions, we can expand
    the matrix element as a perturbative series in
    $\alpha_{s}$ to write
    \begin{equation}\label{eqn:matrix_element}
        |\mathcal{M}|^{2} =  \alpha_{s}^{m} |\mathcal{M}|^{2}_{\mathrm{LO}} + \alpha_{s}^{m+1} |\mathcal{M}|^{2}_{\mathrm{NLO}} + \alpha_{s}^{m+2} |\mathcal{M}|^{2}_{\mathrm{NNLO}} + \mathcal{O}(\alpha_{s}^{m+3}) \, ,
    \end{equation}
    where $m$ corresponds to the powers of $\alpha_{s}$
    in the simplest Feynman diagrams you can draw
    for the process of interest.
    Since each term of this expansion is at a fixed-order
    in $\alpha_{s}$, it is possible to systematically
    compute the diagrams which contribute at the given order of $\alpha_{s}$.
    The set of diagrams that contribute at the lowest
    order in $\alpha_{s}$ are $|\mathcal{M}|^{2}_{\mathrm{LO}}$,
    where LO stands for leading order. The next term
    in the expansion corresponds to the leading order
    diagrams with an additional loop, or external leg,
    which contributes an extra factor of $\alpha_{s}$.
    This term is dubbed NLO for next-to-leading order
    in $\alpha_{s}$, and the following term
    next-to-next-to-leading order has again an additional
    loop or leg. In general, we have terms
    N$^{k}$LO where each additional power of $\alpha_{s}$
    increases complixity of the computations, however,
    because the expansion $\alpha_{s}$
    is small, each additional term contribute less.
    This means that the first terms dominate the expansion,
    and it is sufficient to terminate the series after
    a few terms to reach an acceptable level of accuracy.
    A further discussion of this point will be given
    in Chapter~\ref{chapter:qcd} where we break down
    the computation of each term in (\ref{eqn:matrix_element}).

@@

    With the introduction of matrix elements and
    cross-sections complete, we will move the discussion onto
    Monte Carlo event generators (MCEG). These event
    generators are multipurpose tools that
    incorporate many aspects of the theory discussed
    so far (and to be discussed) for practical use.
    Most importantly, they are
    used to compare theoretical predictions to
    experimental measurements.

>>>>

    \begin{figure}\label{fig:qcd_feynman_rules}
        \includegraphics[width=\linewidth]{introduction/qcd_feynman_rules.pdf}
        \caption{QCD Feynman rules.}
    \end{figure}

    In Figure~\ref{fig:qcd_feynman_rules} we have
    collected the QCD Feynman rules.
    Note that we have only written down the vertices,
    propagators, and external lines for the physical gluons and quarks,
    ghosts also have associated Feynman rules but we do
    not collect those here. A full list of Feynman rules
    in the SM can be found at Ref. \cite{Romao:2012pq}.

@@

    For a review of PDFs and how they are determined
    see Ref. \cite{Ethier:2020way,Jimenez-Delgado:2013sma}.
    In practice, these PDF fits are accessed through the
    LHAPDF interface \cite{Buckley:2014ana} which provides
    PDF sets from multiple working groups \cite{PDF4LHCWorkingGroup:2022cjn,Hou:2019efy,Bailey:2020ooq,NNPDF:2021njg}.
    The evolution of PDFs between factorisation scales is possible
    through the use of the Dokshitzer-Gribov-Lipatov-Altarelli-Parisi (DGLAP)
    equations \cite{Dokshitzer:1977sg,Gribov:1972ri,Lipatov:1974qm,Altarelli:1977zs}.
    \begin{equation}\label{eqn:DGLAP}
        \mu_{F}^{2} \dfrac{\partial}{\partial \mu_{F}^{2}}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \\
        = \dfrac{\alpha_{s}(\mu_{F})}{2\pi} \int_{x}^{1} \dfrac{\mathrm{d}z}{z}
        \begin{pmatrix}
            P_{qq}(\frac{x}{z}) & P_{qg}(\frac{x}{z}) \\
            P_{gq}(\frac{x}{z}) & P_{gg}(\frac{x}{z})
        \end{pmatrix}
        \begin{pmatrix}
            f_{q/h}(x,\mu_{F}) \\
            f_{g/h}(x,\mu_{F})
        \end{pmatrix} \, ,
    \end{equation}
    where $P_{ba}(\dfrac{x}{z})$ are splitting functions
    representing parton $b$ emitting a parton $a$ carrying
    a momentum fraction $z$. These splitting functions are calculable
    as a power series in $\alpha_{s}$ where they have been computed
    up to three-loops \cite{Moch:2004pa,Vogt:2004mw}.
    At leading order \cite{Altarelli:1977zs} they are
    \begin{align}\label{eqn:AP_kernels}
        P^{(0)}_{qq}(z) &= C_{F}\left[\dfrac{1+z^{2}}{(1-z)_{+}} + \dfrac{3}{2}\delta(1-z)\right] \\
        P^{(0)}_{qg}(z) &= T_{R}\left[z^{2} + (1-z)^{2}\right] \\
        P^{(0)}_{gq}(z) &= C_{F}\left[\dfrac{1+(1-z)^{2}}{z^{2}}\right] \\
        P^{(0)}_{gg}(z) &= 2C_{A}\left[\dfrac{z}{(1-z)_{+}}+\dfrac{1-z}{z}+z(1-z)\right] + \left(\dfrac{11C_{A}-4T_{R}n_{f}}{6}\right)\delta(1-z)
    \end{align}
    where the `$+$'-prescription has been used to regulate divergences
    at $z = 1$.
    \begin{equation}\label{eqn:plus_prescription}
        \int_{0}^{1} \mathrm{d}z \; [g(z)]_{+}f(z) = \int_{0}^{1}\mathrm{d}z \; g(z)(f(z)-f(1)) \, ,
    \end{equation}
    where the function $f(z)$ cancels the divergence if it is
    sufficiently smooth at $z=1$.
    At leading order, there is a physical interpretation
    of the splitting kernels as the probability of finding a parton
    $a$ in parton $b$ carrying a fraction $x$ of the parent parton.

@@

    From (\ref{eqn:S_prob}), it becomes clear
    that the matrix element is the object which
    relates back to the Lagrangian of the theory
    as it encodes the interactions of the particles.
    We saw in Section \ref{sec:qcd_feynman} that
    interactions could be codified into Feynman
    rules which are pieced together to construct
    Feynman diagrams. In this picture, matrix elements
    are exactly the sum over all Feynman diagrams
    \footnote{technically the interference of
    Feynman diagrams as we are working on the level
    of $|\mathcal{M}|^{2}$}
    for a particular process. However, there are
    an infinite number of Feynman diagrams you can
    draw as there is no constraints on how the
    vertices and propagators can be connected.
    Fortunately because $\alpha_{s}$
    is small in high energy collisions, we can expand
    the matrix element as a perturbative series in
    $\alpha_{s}$ to write
    \begin{equation}\label{eqn:matrix_element}
        |\mathcal{M}|^{2} =  \alpha_{s}^{m} |\mathcal{M}|^{2}_{\mathrm{LO}} + \alpha_{s}^{m+1} |\mathcal{M}|^{2}_{\mathrm{NLO}} + \alpha_{s}^{m+2} |\mathcal{M}|^{2}_{\mathrm{NNLO}} + \mathcal{O}(\alpha_{s}^{m+3}) \, ,
    \end{equation}
    where $m$ corresponds to the powers of $\alpha_{s}$
    in the simplest Feynman diagrams you can draw
    for the process of interest.
    Since each term of this expansion is at a fixed-order
    in $\alpha_{s}$, it is possible to systematically
    compute the diagrams which contribute at the given order of $\alpha_{s}$.
    The set of diagrams that contribute at the lowest
    order in $\alpha_{s}$ are $|\mathcal{M}|^{2}_{\mathrm{LO}}$,
    where LO stands for leading order. The next term
    in the expansion corresponds to the leading order
    diagrams with an additional loop, or external leg,
    which contributes an extra factor of $\alpha_{s}$.
    This term is dubbed NLO for next-to-leading order
    in $\alpha_{s}$, and the following term
    next-to-next-to-leading order has again an additional
    loop or leg. In general, we have terms
    N$^{\mathrm{k}}$LO where each additional power of $\alpha_{s}$
    increases complixity of the computations, however,
    because the expansion $\alpha_{s}$
    is small, each additional term contribute less.
    This means that the first terms dominate the expansion,
    and it is sufficient to terminate the series after
    a few terms to reach an acceptable level of accuracy.
    A further discussion of this point will be given
    in Chapter~\ref{chapter:qcd} where we break down
    the computation of each term in (\ref{eqn:matrix_element}).

@@

    With the introduction of matrix elements and
    cross-sections complete, we will move the discussion onto
    Monte Carlo event generators. These event
    generators are multipurpose tools that
    incorporate many aspects of the theory discussed
    so far (and to be discussed) for practical use.
    Most importantly, they are
    used to compare theoretical predictions to
    experimental measurements.
